### 2007-01-15 Documentation/Paper2/paper2.tex
*** empty log message ***

\begin{abstract}
We present Version 3.0 of the GLoBES (``General Long Baseline Experiment Simulator'') software, which is
a simulation tool for short- and long-baseline neutrino oscillation experiments. As new features, \GLOBESN\ allows the simulation of experiments with multiple discrete sources and detectors by the 
concept of user-defined systematics.
In addition, the combination with external information, such as from different experiment classes, is 
simplified by the concept of user-defined priors. As far as the probability calculation is concerned, \GLOBES\ 
now provides an interface for the inclusion of non-standard physics without re-compliation of the software. 
The set of experiment descriptions coming with \GLOBES\ has been updated. For example, built-in fluxes are now
provided for the simulation of beta beams.
\end{abstract}

@@

The probability calculation in \GLOBES\ is based on the diagonalization of the
Hamiltonian in layers of constant matter density using the standard three flavor
scenario of neutrino oscillations. In flavor base, we have
\begin{equation}
\mathcal{H}(n_e) = \frac{1}{2E} \, U \,
\left(
\begin{array}{ccc}
0 & 0 & 0 \\
0  & \sdm & 0 \\
0  & 0 & \ldm \\
\end{array}
\right) \,
U^\dagger + 
\left(
\begin{array}{ccc}
 \pm \sqrt{2} G_F n_e & 0 & 0 \\
0 & 0 & 0 \\
0 & 0 & 0 
\end{array}
\right)
\label{equ:ham}
\end{equation}
with $U$ the mixing matrix (described by four parameters $\theta_{12}$, $\theta_{13}$, $\theta_{23}$, and $\deltacp$), and $n_e$ the constant electron density
in that matter density layer. The electron density is related to the matter density $\rho$ by 
$n_e \simeq 0.5 \, \rho/m_N$ with $m_N$ being the nucleon mass. The sign in the second term depends on if one
uses neutrinos or antineutrinos. For $k$ matter density layers with densities
$n^k_e$ and thicknesses $x_i$, the oscillation probability then evaluates to
\begin{equation}
P(\nu_\alpha \rightarrow \nu_\beta) = \left| \langle \nu_\beta | \mathcal{E}(n^k_e, x_k) \hdots  \mathcal{E}(n^1_e, x_1) | \nu_\alpha \rangle \right|^2
\label{equ:probs}
\end{equation}
with the evolution operators (CORRECT REFERENCE?)
\begin{equation}
  \mathcal{E}(n^k_e, x_k) = \exp \left( -i \mathcal{H}(n^k_e) x_k \right) = \frac{1}{2E} \, \hat{U}(n^k_e) \, \left( \begin{array}{ccc}
e^{-i \, m_1^k \, x_k} & 0 & 0 \\
0 &  e^{-i \, m_2^k \, x_k} & 0 \\
0 & 0 & e^{-i \, m_3^k \, x_k} 
\end{array}
\right) \, \hat{U}(n^k_e)^\dagger \, .
\end{equation}
Here the Hamiltonian $\mathcal{H}(n^k_e)$ is diagonalized by the unitary mixing matrix in matter $\hat{U}(n^k_e)$
with the eigenvalues $m_1^k$, $m_2^k$, and $m_3^k$. The probabilities in \equ{probs} are used for the
event rate compution, which is described in greater detail in \Ref~\cite{Huber:2004ka}. In the next section,
we will demonstrate how systematics is implemented thereafter.

@@

\GLOBESN\ allows the modification of this Hamiltonian and the whole probability engine without
re-compilation of the \GLOBES\ software. This is implemented
by using pointers for the functions calling the standard probability engine, which can be changed if
a different probability engine is registered. For example, a user may just copy the standard probability
engine of \GLOBES , and modify it, and register it. An important prerequisite for the ability to handle
more than the standard six oscillation parameters (plus matter density) $\boldsymbol{\lambda}=(\theta_{12},\theta_{13}, \theta_{23}, \deltacp, \sdm, \ldm, \rho)$, \ie,
\begin{equation}
\boldsymbol{\lambda}=(\theta_{12},\theta_{13}, \theta_{23}, \deltacp, \sdm, \ldm, \rho, \eta_1, \hdots, \eta_n)
\end{equation}
with $n$ non-standard parameters $\eta_j$. These parameters can be accessed in the usual way, see \Ref~\cite{Manual}.
For example, non-standard Hamiltonian effects in the $e$-$\tau$-sector may be introduced by modifying \equ{ham} to
\begin{equation}
 \mathcal{\hat{H}} = \mathcal{H}(n_e) + 
\left(
\begin{array}{ccc}
0 & 0 & \epsilon_{e \tau} \\
0 & 0 & 0 \\
\epsilon_{e \tau}^* & 0 & 0 
\end{array}
\right)
\label{equ:nsi}
\end{equation}
In this case, there are two more real parameters, such as the absolute value and phase of $\epsilon_{e \tau}$.
In the literature, the non-standard physics feature in \GLOBES\ as experimental feature has been used for damping effects (such as neutrino decoherence, decay, \etc) in \Ref~\cite{Blennow:2005yk}, Hamiltonian-level effects (such as non-standard matter effects) in \Ref~\cite{Blennow:2005qj}, and mass-varying neutrinos in \Ref~\cite{Schwetz:2005fy} by implementing an environment dependence of neutrino mass.  ANY OTHERS (PH!)?
We show in \figu{reactorNSI} a result from \Ref~\cite{Blennow:2005yk} provided as {\tt example6.c} in the \GLOBES\
distribution. Compared to \equ{nsi}, the underlying physics takes place at the probability level, \ie, 
\equ{probs} (for details, see \Ref~\cite{Blennow:2005yk}). In {\tt example6.c}, an analytical implementation
of this physics is used for simplicity.

@@

\GLOBES\ supports four types of systematical errors as a standard: Signal and background
normalization errors, as well as signal and background tilt or energy calibration errors.
The tilt (T) is impemented as a linear distortion of the spectrum around the center,
and the energy calibration (C) as a distortion (stretching) of the reconstructed energy scale.
For the implementation of these systematical errors, the  ``pull method'' is used~\cite{Fogli:2002pt}, which introduces  so-called nuisance parameters $\zeta_i$. 
For example, for the signal and background normalization errors, the respective event rates 
$s_i$ and $b_i$ in
each bin $i$ are multiplied with\footnote{For details on the event rate calculation, see \Ref~\cite{Huber:2004ka}.}
\begin{equation}
s_i(\zeta_1) := (1+\zeta_1)\cdot s_i \, , \quad b_i(\zeta_2) : =(1+\zeta_2)\cdot b_i \, ,
\end{equation} 
\ie, the rates are scaled by these parameters. The systematics $\chi^2$ is then minimized over
these parameters:
\begin{equation}
\chi^2_\mathrm{pull}(\boldsymbol{\lambda}):=\min_{\{\zeta_i\} } \,\, \left( \,
\chi^2(\boldsymbol{\lambda},
\zeta_1, \ldots, \zeta_k)+ \sum_{j=1}^{k} \frac{\zeta_j^2}{\sigma_{\zeta_j}^2}
\, \right)\,.
\label{equ:chipull}
\end{equation}
Here $\chi^2(\boldsymbol{\lambda}, \zeta_1, \ldots, \zeta_k)$ is the usual Possonian $\chi^2$
depending on the neutrino oscillation parameters $\boldsymbol{\lambda}$ and the nuisance
parameters $\zeta_i$ (which, for instance, scale the signal or background rates).
In addition, Gaussian penalties $\zeta_j^2/\sigma_{\zeta_j}^2$ are added, where 
$\sigma_{\zeta_j}$ corresponds to the actual systematical error.
Core part of \equ{chipull} is the definition of $\chi^2$. For example, for a background-free measurement and
a signal normalization error only, it reads
\begin{equation}
  \chi^2 (\boldsymbol{\lambda},a) =  \sum_{i=1}^{\textrm{\# of bins}}2
     \left( O_{i} - (1 + a) T_{i} +
       (1 + a) T_{i} \log \frac{(1 + a) T_{i}}{O_{i}}  \right) + \frac{a^2}{\sigma_a^2} 
\label{equ:chistandard}
\end{equation}
(JK/PH: CHECK THIS EQUATION IF THIS IS THE ACTUAL IMPLEMENTATION. HERE THE SYSTEMATICS IS\\
IMPLEMENTED IN THE SIMULATED RATES. IN ADDITION, CHECK IF THEORETICAL AND EXP. RATES
ARE CORRECTLY GIVEN. [WW])

@@

In order to include correlations and degeneracies, \GLOBES\ provides
(local) maginalization routines to project the fit manifold onto
a subspace. For example, if one wants to compute the $\stheta$ precision,
one scans the $\stheta$ direction and marginalizes for each fixed $\stheta$
over the other oscillation parameters. This marginalization is performed
after the systematics $\chi^2_{\mathrm{pull}}$ in \equ{chipull} (or similar) 
has been determined:
\begin{equation}
 \chi^2_{\mathrm{proj}}(\boldsymbol{\lambda_1}) := \min_{\{ \boldsymbol{\lambda_2} \} } \left( \, \chi^2_{\mathrm{pull}} (\boldsymbol{\lambda_1} , \boldsymbol{\lambda_2} ) \, \right) \, .
\label{equ:margin}
\end{equation}
In this case, the parameter space of oscillation parameters 
$\boldsymbol{\lambda}$ is projected onto the subspace $\boldsymbol{\lambda_1}$.
If one wants to include external input which only depends on the oscillation
parameters and not on systematics, one can do this at this level. \GLOBES\ provides
the pre-defined possibility to include external Gaussian constraints on the 
oscillation parameters (and the matter density, which is treated as another
oscillation parameter). For example, an external constraint on $\theta_{12}$
is added before the marginalition in \equ{margin} by the replacement after the
systematics minimization
\begin{equation}
 \chi^2_{\mathrm{pull}}(\boldsymbol{\lambda}) \rightarrow \chi^2_{\mathrm{pull}}(\boldsymbol{\lambda})  +
 \frac{(\theta_{12} - \theta_{12}^0)^2}{\sigma_{\theta_{12}}^2}
\label{equ:priors}
\end{equation}
with the central value $\theta_{12}^0$, where the input is added, and
the $1\sigma$ Gaussian error $\sigma_{\theta_{12}}$. This corresponds to an external
measurement of the solar mixing angle at the best-fit value $\theta_{12}^0$
with the error $\sigma_{\theta_{12}}$.

@@

\AEDL\ (Abstract Experiment Definition Language) is a special syntax developed for the description of
experiments in \GLOBES\ represented by simple text files. Core part of \AEDL\ are the following constructions
(for details, see \Ref~\cite{Manual}):
\begin{description}
\item[Channel] A channel represents an oscillation channel from the source flux, over the initial and final
flavors and the use of neutrinos or antineutrinos, to the interaction type and cross section. The result
is an event rate vector.
\item[Rule] A rule combines one or more signal channels with one or more background channels. The rates
from these channels are added, and the signal and background are implemented with specific systematics, \ie,
systematics is connected to the concept of a rule.
The result of a rule is a systematics $\chi^2$ corresponding to \equ{chipull}.
\item[Experiment] An experiment is a combination of one or more rules, which may correspond to different
appearance and disappearance channels, neutrino and antineutrino operations modes, \etc. An experiment has a common detector and baseline, but may use more different fluxes (such as for neutrinos and antineutrinos). Therefore,
the matter density profile is the same for all rules of an experiment. The result of an experiment
is again a systematics $\chi^2$, but summed over all rules.
\item[Experiment combination] An experiment combination consists of different experiments with different detectors, baselines, and (uncorreleted) matter density profiles. The result of an experiment combination
is again a systematics $\chi^2$, but summed over all experiments.
\end{description}
Note that after the $\chi^2$ is summed over all rules or experiments as specified by the user in the application software, \equ{margin} is applied, and the oscillation parameters are marginalized over.

In \GLOBESN , this construction of experiments is extended by user-defined systematics, which allows
the user to break up the strict coupling of systematics with rules. For example, systematics can be
correlated between different rules or experiments, such as for an reactor experiment with identical
near and far detectors. In addition, one can introduce user-defined systematical errors.
For that purpose, the former error dimension concept has been changed and
extended. For example, one has to give each user-defined systematics a name, and this name has
to be matched by the application software. 
%
Further changes in \AEDL\ are the requirement of a minimum \GLOBES\ version the \AEDL\ is running with,
and the extension of lists as \AEDL\ variables and an interpolation feature.

As far as the experiment prototypes are concerned, a list of the ones provided with \GLOBESN\ 
is given in \Tab~\ref{tab:experiments}. These prototypes can be modified for the definition of
one's own experiments (experimentalist), or they can be used for the test of new physics concepts
(phenomenologist/theorist).
Note that some of these experiments may contain 
integrated luminosities, baseline, fluxes, efficiencies, or other factors, which the user
may not agree with and which can be easily adjusted. Some of the \AEDL\ files have been
changed or updated from eariler versions, such as for T2K and T2HK, or the standard neutrino
factory. Other files represent the results from recent developments, such as the variable energy
beta beam or neutrino factory files. They push \AEDL\ to the edge in terms of advanced features,
such as variable bin widths, use of the new interpolation feature, threshold function implementations, \etc.
In particular, note that Double Chooz requires user-defined
systematics, and the beta beams built-in beta beam fluxes. Both of these features are only
available in \GLOBESN\ and higher. Some of the files require that \AEDL\ variables be pre-set by the
user, such as the muon energy for the variable energy neutrino factory. These variables are
described in the corresponding file headers. Old \AEDL\ prototypes should still run, but will not
be maintained by the \GLOBES\ team anymore.

@@

New platform and interpreter support, new ways to improve the performance

\bi
\item
 PERL?
\item 
 Windows?
\item
 Better RedHat support
\item
 Changes in installation scheme
\ei

\bi
\item 
 GLoBES on clusters, Condor; parallized code; static links
\item 
 64 bit support
\item 
 New minimization algorithm!?
\ei

@@

Firstly, the \GLOBES\ minimizers distinguish between minimization over
oscillation parameters and minimization over systematics parameters.
This is necessary because each step in the oscillation parameter space
requires a time-consuming re-computation of oscillation probabilities,
while a modification of the systematics parameters does not. The standard
procedure is to perform a full minimization in systematics space after each
step in oscillation space. This avoids interference of the two categories
of parameters, and is known to have excellent convergence properties.
The numerical method that does the minimization is the Powell
algorithm~\cite{Press:NumRecip}.

In \GLOBESN , we provide an alternative method, which does the full minimization
in one go and is therefore several times faster in most cases. In this algorithm,
each Powell iteration over oscillation parameters is followed by one iteration
over systematics parameters. Since there may be some pathological cases where this
interleaved minimization strategy behaves differently than the old method,
it is by default not used in \GLOBESN . This ensures maximal compatibility
with old application programs, but it requires users who wish to benefit from the
new algorithm to explicitly select it by an API function call.

Besides optimizing the minimization functions, it is also mandatory to
optimize the routines which are called by them to calculate $\chi^2$
for one specific set of parameters. In particular, the computation of
oscillation probabilities, which has to be done every time the oscillation
parameters have changed, should be as efficient as possible. The most expensive
step here is the diagonalization of the hermitian $3\times3$ neutrino Hamilton
operator in matter. The LAPACK rotuines~\cite{Anderson:LAPACK}, which were
used to solve this eigenproblem in previous versions of \GLOBES, turned out
not to be the optimal choice because they are mainly optimized for very large
matrices~\cite{Kopp:2006wp}. In the case of the small and usually well-conditioned
matrices appearing in \GLOBES, they spend too much time evaluating their parameter
list and planning the optimal diagonalization strategy. Therefore, \GLOBES\ now
relies an algorithm which has been designed specifically for the diagonalization
of hermitian $3\times3$ matrices. It is essentially the extremely fast analytical
algorithm discussed in~\cite{Kopp:2006wp}, but for the rare cases where this may become
inaccurate, it falls back to the QL method from~\cite{Press:NumRecip}.

@@

>>>>

\begin{abstract}
We present Version 3.0 of the GLoBES (``General Long Baseline Experiment Simulator'') software, which is
a simulation tool for short- and long-baseline neutrino oscillation experiments. As new features, \GLOBESN\ allows the simulation of experiments with multiple discrete sources and detectors by the 
concept of user-defined systematics.
In addition, the combination with external information, such as from different experiment classes, is 
simplified by the concept of user-defined priors. As far as the probability calculation is concerned, \GLOBES\ 
now provides an interface for the inclusion of non-standard physics without re-compilation of the software. 
The set of experiment descriptions coming with \GLOBES\ has been updated. For example, built-in fluxes are now
provided for the simulation of beta beams.
\end{abstract}

@@

The probability calculation in \GLOBES\ is based on the diagonalization of the
Hamiltonian in layers of constant matter density using the standard three flavor
scenario of neutrino oscillations. In flavor base, we have
\begin{equation}
\mathcal{H}(n_e) = \frac{1}{2E} \, U \,
\left(
\begin{array}{ccc}
0 & 0 & 0 \\
0  & \sdm & 0 \\
0  & 0 & \ldm \\
\end{array}
\right) \,
U^\dagger + 
\left(
\begin{array}{ccc}
 \pm \sqrt{2} G_F n_e & 0 & 0 \\
0 & 0 & 0 \\
0 & 0 & 0 
\end{array}
\right)
\label{equ:ham}
\end{equation}
with $U$ the mixing matrix (described by four parameters $\theta_{12}$, $\theta_{13}$, $\theta_{23}$, and $\deltacp$), and $n_e$ the constant electron density
in that matter density layer. The electron density is related to the matter density $\rho$ by 
$n_e \simeq 0.5 \, \rho/m_N$ with $m_N$ being the nucleon mass. The sign in the second term depends on if one
uses neutrinos or antineutrinos. For $k$ matter density layers with densities
$n^k_e$ and thicknesses $x_i$, the oscillation probability then evaluates to
\begin{equation}
P(\nu_\alpha \rightarrow \nu_\beta) = \left| \langle \nu_\beta | \mathcal{E}(n^k_e, x_k) \hdots  \mathcal{E}(n^1_e, x_1) | \nu_\alpha \rangle \right|^2
\label{equ:probs}
\end{equation}
with the evolution operators (CORRECT REFERENCE?)
\begin{equation}
  \mathcal{E}(n^k_e, x_k) = \exp \left( -i \mathcal{H}(n^k_e) x_k \right) = \frac{1}{2E} \, \hat{U}(n^k_e) \, \left( \begin{array}{ccc}
e^{-i \, m_1^k \, x_k} & 0 & 0 \\
0 &  e^{-i \, m_2^k \, x_k} & 0 \\
0 & 0 & e^{-i \, m_3^k \, x_k} 
\end{array}
\right) \, \hat{U}(n^k_e)^\dagger \, .
\end{equation}
Here the Hamiltonian $\mathcal{H}(n^k_e)$ is diagonalized by the unitary mixing matrix in matter $\hat{U}(n^k_e)$
with the eigenvalues $m_1^k$, $m_2^k$, and $m_3^k$. The probabilities in \equ{probs} are used for the
event rate computation, which is described in greater detail in \Ref~\cite{Huber:2004ka}. In the next section,
we will demonstrate how systematics is implemented thereafter.

@@

\GLOBESN\ allows the modification of this Hamiltonian and the whole probability engine without
re-compilation of the \GLOBES\ software. This is implemented
by using pointers to the functions of the standard probability engine, which can be changed if
a different probability engine is registered. For example, a user may just copy the standard probability
engine of \GLOBES , modify it, and register it. An important prerequisite for the ability to handle
more than the standard six oscillation parameters (plus matter density) $\boldsymbol{\lambda}=(\theta_{12},\theta_{13}, \theta_{23}, \deltacp, \sdm, \ldm, \rho)$, \ie,
\begin{equation}
\boldsymbol{\lambda}=(\theta_{12},\theta_{13}, \theta_{23}, \deltacp, \sdm, \ldm, \rho, \eta_1, \hdots, \eta_n)
\end{equation}
with $n$ non-standard parameters $\eta_j$. These parameters can be accessed in the usual way, see \Ref~\cite{Manual}.
For example, non-standard Hamiltonian effects in the $e$-$\tau$-sector may be introduced by modifying \equ{ham} to
\begin{equation}
 \mathcal{\hat{H}} = \mathcal{H}(n_e) + 
\left(
\begin{array}{ccc}
0 & 0 & \epsilon_{e \tau} \\
0 & 0 & 0 \\
\epsilon_{e \tau}^* & 0 & 0 
\end{array}
\right)
\label{equ:nsi}
\end{equation}
In this case, there are two more real parameters, such as the absolute value and phase of $\epsilon_{e \tau}$.
In the literature, the non-standard physics feature in \GLOBES\ as experimental feature has been used for damping effects (such as neutrino decoherence, decay, \etc) in \Ref~\cite{Blennow:2005yk}, Hamiltonian-level effects (such as non-standard matter effects) in \Ref~\cite{Blennow:2005qj}, and mass-varying neutrinos in \Ref~\cite{Schwetz:2005fy} by implementing an environment dependence of neutrino mass.  ANY OTHERS (PH!)?
We show in \figu{reactorNSI} a result from \Ref~\cite{Blennow:2005yk} provided as {\tt example6.c} in the \GLOBES\
distribution. Compared to \equ{nsi}, the underlying physics takes place at the probability level, \ie, 
\equ{probs} (for details, see \Ref~\cite{Blennow:2005yk}). In {\tt example6.c}, an analytical implementation
of this physics is used for simplicity.

@@

\GLOBES\ supports four types of systematical errors as a standard: Signal and background
normalization errors, as well as signal and background tilt or energy calibration errors.
The tilt (T) is implemented as a linear distortion of the spectrum around the center,
and the energy calibration (C) as a distortion (stretching) of the reconstructed energy scale.
For the implementation of these systematical errors, the  ``pull method'' is used~\cite{Fogli:2002pt}, which introduces  so-called nuisance parameters $\zeta_i$. 
For example, for the signal and background normalization errors, the respective event rates 
$s_i$ and $b_i$ in
each bin $i$ are multiplied with\footnote{For details on the event rate calculation, see \Ref~\cite{Huber:2004ka}.}
\begin{equation}
s_i(\zeta_1) := (1+\zeta_1)\cdot s_i \, , \quad b_i(\zeta_2) : =(1+\zeta_2)\cdot b_i \, ,
\end{equation} 
\ie, the rates are scaled by these parameters. The systematics $\chi^2$ is then minimized over
these parameters:
\begin{equation}
\chi^2_\mathrm{pull}(\boldsymbol{\lambda}):=\min_{\{\zeta_i\} } \,\, \left( \,
\chi^2(\boldsymbol{\lambda},
\zeta_1, \ldots, \zeta_k)+ \sum_{j=1}^{k} \frac{\zeta_j^2}{\sigma_{\zeta_j}^2}
\, \right)\,.
\label{equ:chipull}
\end{equation}
Here $\chi^2(\boldsymbol{\lambda}, \zeta_1, \ldots, \zeta_k)$ is the usual Possonian $\chi^2$
depending on the neutrino oscillation parameters $\boldsymbol{\lambda}$ and the nuisance
parameters $\zeta_i$ (which, for instance, scale the signal or background rates).
In addition, Gaussian penalties $\zeta_j^2/\sigma_{\zeta_j}^2$ are added, where 
$\sigma_{\zeta_j}$ corresponds to the actual systematical error.
Core part of \equ{chipull} is the definition of $\chi^2$. For example, for a background-free measurement and
a signal normalization error only, it reads
\begin{equation}
  \chi^2 (\boldsymbol{\lambda},a) =  \sum_{i=1}^{\textrm{\# of bins}}2
     \left( O_{i} - (1 + a) T_{i} +
       (1 + a) T_{i} \log \frac{(1 + a) T_{i}}{O_{i}}  \right) + \frac{a^2}{\sigma_a^2} 
\label{equ:chistandard}
\end{equation}
(JK/PH: CHECK THIS EQUATION IF THIS IS THE ACTUAL IMPLEMENTATION. HERE THE SYSTEMATICS IS\\
IMPLEMENTED IN THE SIMULATED RATES. IN ADDITION, CHECK IF THEORETICAL AND EXP. RATES
ARE CORRECTLY GIVEN. [WW])

@@

In order to include correlations and degeneracies, \GLOBES\ provides
(local) marginalization routines to project the fit manifold onto
a subspace. For example, if one wants to compute the $\stheta$ precision,
one scans the $\stheta$ direction and marginalizes for each fixed $\stheta$
over the other oscillation parameters. This marginalization is performed
after the systematics $\chi^2_{\mathrm{pull}}$ in \equ{chipull} (or similar) 
has been determined:
\begin{equation}
 \chi^2_{\mathrm{proj}}(\boldsymbol{\lambda_1}) := \min_{\{ \boldsymbol{\lambda_2} \} } \left( \, \chi^2_{\mathrm{pull}} (\boldsymbol{\lambda_1} , \boldsymbol{\lambda_2} ) \, \right) \, .
\label{equ:margin}
\end{equation}
In this case, the parameter space of oscillation parameters 
$\boldsymbol{\lambda}$ is projected onto the subspace $\boldsymbol{\lambda_1}$.
If one wants to include external input which only depends on the oscillation
parameters and not on systematics, one can do this at this level. \GLOBES\ provides
the pre-defined possibility to include external Gaussian constraints on the 
oscillation parameters (and the matter density, which is treated as another
oscillation parameter). For example, an external constraint on $\theta_{12}$
is added before the marginalization in \equ{margin} by the replacement
\begin{equation}
 \chi^2_{\mathrm{pull}}(\boldsymbol{\lambda}) \rightarrow \chi^2_{\mathrm{pull}}(\boldsymbol{\lambda})  +
 \frac{(\theta_{12} - \theta_{12}^0)^2}{\sigma_{\theta_{12}}^2}
\label{equ:priors}
\end{equation}
with the central value $\theta_{12}^0$, where the input is added, and
the $1\sigma$ Gaussian error $\sigma_{\theta_{12}}$. This corresponds to an external
measurement of the solar mixing angle at the best-fit value $\theta_{12}^0$
with the error $\sigma_{\theta_{12}}$.

@@

\AEDL\ (Abstract Experiment Definition Language) is a special syntax developed for the description of
experiments in \GLOBES\ represented by simple text files. Core part of \AEDL\ are the following constructions
(for details, see \Ref~\cite{Manual}):
\begin{description}
\item[Channel] A channel represents an oscillation channel from the source flux, over the initial and final
flavors and the use of neutrinos or antineutrinos, to the interaction type and cross section. The result
is an event rate vector.
\item[Rule] A rule combines one or more signal channels with one or more background channels. The rates
from these channels are added, and the signal and background are implemented with specific systematics, \ie,
systematics is connected to the concept of a rule.
The result of a rule is a systematics $\chi^2$ corresponding to \equ{chipull}.
\item[Experiment] An experiment is a combination of one or more rules, which may correspond to different
appearance and disappearance channels, neutrino and antineutrino operations modes, \etc. An experiment has a common detector and baseline, but may use more different fluxes (such as for neutrinos and antineutrinos). Therefore,
the matter density profile is the same for all rules of an experiment. The result of an experiment
is again a systematics $\chi^2$, but summed over all rules.
\item[Experiment combination] An experiment combination consists of different experiments with different detectors, baselines, and (uncorrelated) matter density profiles. The result of an experiment combination
is again a systematics $\chi^2$, but summed over all experiments.
\end{description}
Note that after the $\chi^2$ is summed over all rules or experiments as specified by the user in the application software, \equ{margin} is applied, and the oscillation parameters are marginalized over.

In \GLOBESN , this construction of experiments is extended by user-defined systematics, which allows
the user to break up the strict coupling between systematics and rules. For example, systematics can be
correlated between different rules or experiments, such as for an reactor experiment with identical
near and far detectors. In addition, one can introduce user-defined systematical errors.
For that purpose, the former error dimension concept has been changed and
extended. For example, one has to give each user-defined systematics a name, and this name has
to be matched by the application software. 
%
Further changes in \AEDL\ are the requirement to specify the minimum \GLOBES\ version the \AEDL\ is running with,
and the extensions by lists as \AEDL\ variables and by an interpolation feature.

As far as the experiment prototypes are concerned, a list of the ones provided with \GLOBESN\ 
is given in \Tab~\ref{tab:experiments}. These prototypes can be modified for the definition of
one's own experiments (experimentalist), or they can be used for the test of new physics concepts
(phenomenologist/theorist).
Note that some of these experiments may contain 
integrated luminosities, baseline, fluxes, efficiencies, or other factors, which the user
may not agree with and which can be easily adjusted. Some of the \AEDL\ files have been
changed or updated from earlier versions, such as for T2K and T2HK, or the standard neutrino
factory. Other files represent the results from recent developments, such as the variable energy
beta beam or neutrino factory files. They push \AEDL\ to the edge in terms of advanced features,
such as variable bin widths, use of the new interpolation feature, threshold function implementations, \etc.
In particular, note that Double Chooz requires user-defined
systematics, and the beta beams require built-in beta beam fluxes. Both of these features are only
available in \GLOBESN\ and higher. Some of the files require that \AEDL\ variables be pre-set by the
user, such as the muon energy for the variable energy neutrino factory. These variables are
described in the corresponding file headers. Old \AEDL\ prototypes should still run, but will not
be maintained by the \GLOBES\ team anymore.

@@

@@

Firstly, the \GLOBES\ minimizers distinguish between minimization over
oscillation parameters and minimization over systematics parameters.
This is necessary because each step in the oscillation parameter space
requires a time-consuming re-computation of oscillation probabilities,
while a modification of the systematics parameters does not. The standard
procedure is to perform a full minimization in systematics space after each
step in oscillation space. This avoids interference of the two categories
of parameters, and is known to have excellent convergence properties.
The numerical method that does the minimization is the Powell
algorithm HERE ORIGINAL POWELL REFERENCE [WW].

In \GLOBESN , we provide an alternative method, which does the full minimization
in one go, and is therefore several times faster in most cases. In this algorithm,
each Powell iteration over oscillation parameters is followed by one iteration
over systematics parameters. Since there may be some pathological cases where this
interleaved minimization strategy behaves differently than the old method,
it is by default not used in \GLOBESN . This ensures maximal compatibility
with old application programs, but it requires users who wish to benefit from the
new algorithm to explicitly select it by an API function call.

Besides optimizing the minimization functions, it is also mandatory to
optimize the routines which are called by them to calculate $\chi^2$
for one specific set of parameters. In particular, the computation of
oscillation probabilities, which has to be done every time the oscillation
parameters have changed, should be as efficient as possible. The most expensive
step here is the diagonalization of the hermitian $3\times3$ neutrino Hamilton
operator in matter. The LAPACK routines~\cite{Anderson:LAPACK}, which were
used to solve this eigenproblem in previous versions of \GLOBES, turned out
not to be the optimal choice because they are mainly optimized for very large
matrices~\cite{Kopp:2006wp}. In the case of the small and usually well-conditioned
matrices appearing in \GLOBES, they spend too much time evaluating their parameter
list and planning the optimal diagonalization strategy. Therefore, \GLOBES\ now
relies an algorithm which has been designed specifically for the diagonalization
of hermitian $3\times3$ matrices. It is essentially the extremely fast analytical
algorithm discussed in \Ref~\cite{Kopp:2006wp}, but for the rare cases where this may become
inaccurate, it falls back to the QL (WRITE OUT!)
method from~\cite{Press:NumRecip} REPLACE BY ORIGINAL REFERENCE!.

As far as the installation of \GLOBES\ on different systems is concerned, \GLOBES\
has been improved in the past, such as to handle different Linux distributions.
In addition, for high-throughput computing, \GLOBES\ has been used or tested
on 64 Bit systems, Condor clusters, and parallel clusters. An important prerequisite
for the parallelization is the option to build static binaries, which is now
illustrated in the examples. With these improvements, the performance of \GLOBES\
has been pushed to the edge. \Ref~\cite{Huber:2006wb} is probably the currently
most demanding product of \GLOBES , for which most of these features
have been developed.

PH, JK: IS THERE ANYTHING ELSE YOU WANT TO SAY? ABOUT NEW PLATFORMS,
PERL, WINDOWS, PERFORMANCE, 64BIT, INSTALLATION?

@@

We have presented \GLOBESN , which is a simulation software for 
future terrestrial neutrino oscillation facilities. \GLOBES\ is now no further
restricted to one source-one detector configurations, but it can simulate
the systematics of arbitrary combinations of discrete sets of sources and
detectors. In addition, user-defined systematical errors can be introduced.
The new concept of user-defined priors allows the inclusion of external
information at a different level. For example, constraints from atmospheric
neutrino experiments or neutrino telescopes can be taken into account. In addition,
they provide a powerful handle on complicated degeneracy localization.
On the probability level, \GLOBES\ now supports the simulation of non-standard
physics beyond three-flavor neutrino oscillations without a re-compilation of
the software. A very important prerequisite for this feature is the ability
of \GLOBESN\ to carry more than six oscillation parameters. New experiment
prototypes make extensive use these new features, as well as they are illustrated by
examples in the manual. One prominent example is the Double Chooz experiment,
which requires user-defined systematics to simulate correlated errors between
the near and far detectors.

In conclusion, \GLOBESN\ provides a number of new improvements, such as by
 including more flexibility. At the end, the developers of a software can never
exactly foresee what the users are up for, and therefore flexible concepts
allow for more space in problem realizations. In this version of \GLOBES ,
three new such flexible concepts have been introduced: User-defined systematics,
user-defined priors, and user-defined oscillation probabilities. These approaches
will make an easy adjustment of the software possible for both experimentalists and theorists.
Finally, note that the past has shown that it is often more than one possibility 
how a specific problem can be implemented in \GLOBES .
